flowCore: data structures package for flow cytometry data

### 2008-10-29 inst/doc/HowTo-flowCore.Rnw
updated filterSummary class and available methods
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33537 bc3139a8-67e5-0310-9ffc-ced21a209358
A first draft of the new workflow infrastructure and all the little changes that were necessary to make that work. This might eventually move to some other place. Also updated the documentation a bit. More to come soon...
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33791 bc3139a8-67e5-0310-9ffc-ced21a209358
added workFlow to vignette
git-svn-id: file:///home/git/hedgehog.fhcrc.

\noindent \textbf{Results} We hope that those proposed data structures
will the base for the development of many tools for the analysis of
high throughput flowcytometry.\\

\noindent \textbf{keywords} Flow cytometry, high throughtput,
software, standard
\end{abstract}

@@

\section{Introduction}
Traditionally, flow cytometry has been a tube-based technique limited
to small-scale laboratory and clinical studies.  High throughput
methods for flow cytometry have recently been developed for drug
discovery and advanced research methods \citep{Gasparetto2004}. As an
example, the flow cytometry high content screening (FC-HCS) can
process up to a thousand samples daily at a single workstation, and
the results have been equivalent or superior to traditional manual
multiparameter staining and analysis techniques.

@@

The aim is to merge both prada and rflowcyt \citep{LeMeur2006} into
one core package wich is compliant with the data exchange standards
that are currently developed in the community \citep{Spidlen2006}.

@@

Finally, only a sample of lines can be read in case you need a quick
overview of a large serie of files.
<<RedFiles3, echo=true, results=verbatim>>= 
lines <- sample(100:500, 50)
y <- read.FCS(file.name, which.lines = lines) 
y 
@

@@

<<rectGate, echo=true, results=verbatim>>=
rectGate <- rectangleGate(filterId="Fluorescence Region", 
                          "FL1-H"=c(50, 100), "FL2-H"=c(50, 100))
@

@@

Though they are not explicitly filters, meaning that they cannot play
a role in the construction of subfilters via operators,
\Rclass{filterSet} objects do define \Rfunction{filter} and
\Rfunction{split} methods that allow for the two most common use
cases. Additionally, a filter can be selected from the
\Rclass{filterSet} using ``[['' to perform a \Rfunction{Subset}
operation if desired.

For standard filtering operations, the \Rclass{filterSet} can yield
better performance than a manual filtering operation because it
ensures that each parent filter is only evaluated once per call to
\Rfunction{filter} so that if many subfilters rely on a particular,
likely expensive, filtering operation they use the original result
rather than recalculating each time. The individual
\Rclass{filterResult} objects may be later extracted for further
analysis of desire, making the \Rclass{filterSet} the logical choice
for most filtering operations. Additionally, though not presently
implemented, it can allow visualization methods to make more
intelligent choices about display when rendering gating strategies
because the disposition of the entire filtering graph is known and the
filter results are already available.

@@

split(fs[[1]],fset1,flowSet=TRUE)

@@

>>>>

\noindent \textbf{Results} We hope that those proposed data structures
will the base for the development of many tools for the analysis of
high throughput flow cytometry.\\

\noindent \textbf{keywords} Flow cytometry, high throughput,
software, standard
\end{abstract}

@@

\section{Introduction}
Traditionally, flow cytometry has been a tube-based technique limited
to small-scale laboratory and clinical studies.  High throughput
methods for flow cytometry have recently been developed for drug
discovery and advanced research methods \citep{Gasparetto2004}. As an
example, the flow cytometry high content screening (FC-HCS) can
process up to a thousand samples daily at a single workstation, and
the results have been equivalent or superior to traditional manual
multi-parameter staining and analysis techniques.

@@

The aim is to merge both prada and rflowcyt \citep{LeMeur2006} into
one core package which is compliant with the data exchange standards
that are currently developed in the community \citep{Spidlen2006}.

@@

Finally, only a sample of lines can be read in case you need a quick
overview of a large series of files.
<<RedFiles3, echo=true, results=verbatim>>= 
lines <- sample(100:500, 50)
y <- read.FCS(file.name, which.lines = lines) 
y 
@

@@

<<rectGate, echo=true, results=verbatim>>=
rectGate <- rectangleGate(filterId="Fluorescence Region", 
                          "FL1-H"=c(0, 12), "FL2-H"=c(0, 12))
@

@@

Though they are not explicitly filters, meaning that they cannot play
a role in the construction of sub filters via operators,
\Rclass{filterSet} objects do define \Rfunction{filter} and
\Rfunction{split} methods that allow for the two most common use
cases. Additionally, a filter can be selected from the
\Rclass{filterSet} using ``[['' to perform a \Rfunction{Subset}
operation if desired.

For standard filtering operations, the \Rclass{filterSet} can yield
better performance than a manual filtering operation because it
ensures that each parent filter is only evaluated once per call to
\Rfunction{filter} so that if many sub filters rely on a particular,
likely expensive, filtering operation they use the original result
rather than recalculating each time. The individual
\Rclass{filterResult} objects may be later extracted for further
analysis of desire, making the \Rclass{filterSet} the logical choice
for most filtering operations. Additionally, though not presently
implemented, it can allow visualization methods to make more
intelligent choices about display when rendering gating strategies
because the disposition of the entire filtering graph is known and the
filter results are already available.

@@

split(fs[[1]], fset1, flowSet=TRUE)

@@

\section{Work flows}
\Rclass{filterSets} are very limited in their use for complex
analysis work flows. They are  result-centric and it is hard to access
intermediate results. \Rpackage{flowCore} offers much more versatile
tools for such tasks though the \Rclass{workFlow} class. The general
idea is to let the software handle the organization of intermediate
results, naming schemes and operations and to provide a unified API
to access and summarize these operations.

\subsection{Abstraction of work flows}
There are three classes in \Rpackage{flowCore} that are used to
abstract work flows: \Rclass{workFlow} objects are the basic container
holding all the necessary bits and pieces and they are the main
structure for user interaction. \Rclass{actionItem} objects are
abstractions of data analysis operations like gating, transformation
or normalization. There are sub-classes for different types of
operations. Finally, \Rclass{view} objects are created when applying
\Rclass{actionItems} to a \Rclass{workFlow}. One can think of
\Rclass{views} as separate data sets, for instance a subset that is
created by a gating operation or the modified data following a
transformation. Again, different types of \Rclass{views} are available
through various sub-classes. This structure allows for a unified API
in which the user interacts with either \Rclass{views} (the most
common application) or \Rclass{actionItems}.

It is important to know that \Rclass{workFlows} use a reference
semantic instead of the pass-by-value semantic that is usually found
in the R language. This design was chosen to minimize memory
consumption and to keep the framework as flexible as possible. The
only main consequence on the user-level is the fact that direct
assignments to a \Rclass{workFlow} object are usually not necessary;
i.e., functions that operate on the \Rclass{workFlow} have the
potential side-effect of modifying the object.

\subsection{Naming schemes}
One major obstacle of using scripts as a work flow representation is
the need to keep track of object symbols or names of list
items. Choosing arbitrary names makes it hard to access objects and
there is no way to easily query the structure of the work flow other
than reading through the complete script. It is also very hard to
appreciate the hierarchical structure of the work flow from a linear
listing of R commands. Most objects in \Rpackage{flowCore} provide
useful names or identifiers, and the work flow framework tries to
utilize these names as much as possible. For example, a
\Rclass{rectangleGate} called ``Lymphocytes'' will create two
subpopulations, namely the cells within the gate and their
complement. Intuitive names for these subpopulation would therefore be
``Lymphocyte+'' and ``Lymphocyte-''. These names don't necessarily
need to be unique, and in order to be able to unambiguously address
each item in a work flow we need additional unique identifiers. Our
software keeps an alias table that maps these internal unique
identifiers to more human-readable and user friendly names, and unless
there are naming collisions, those names will also be the primary
identifiers that are exposed in the API. Only if and alias is not
unique, objects need to be addressed by their internal unique
identifier.

\subsection{Creating \Rclass{workFlow} objects}
Objects of class \Rclass{workFlow} have to be created using the
constructor \Rfunction{workFlow}. The single mandatory argument is a
flow data object, either a \Rclass{flowFrame} or a \Rclass{flowSet}.
<<createWorkFlow>>=
data(GvHD)
wf <- workFlow(GvHD[1:5], name="myWorkflow")
wf
@
%
Printing the objects shows its structure: the work flow contains a
single \Rclass{view} ``base view'' which essentially is the
\Rclass{flowSet} \Robject{fs}. Since the data was added by the
constructor and is not the result of a particular data analysis
operation, no \Rclass{actionItem} is associated to the
\Rclass{view}. We can query the available views in the
\Rclass{workFlow} using the \Rfunction{views} method:
<<views>>=
views(wf)
@

\subsection{Adding  \Rclass{actionItems}}
Currently there are four different types of \Rclass{actionItems} that
can be added to a \Rclass{workFlow}:

\begin{itemize}
\item{\Rclass{compensateActionItem}: A \Rclass{compensation} object}
\item{\Rclass{transformActionItem}: A \Rclass{transformList} object}
\item{\Rclass{normalizeActionItem}: A \Rclass{normalization} object}
\item{\Rclass{gateActionItem}: A \Rclass{filter} object}
\end{itemize}

The user doesn't have to bother with the details of these classes,
they will be created internally when adding one of the above source
objects to the \Rclass{workFlow} using the \Rfunction{add} method.
<<addItems>>=
tf <- transformList(colnames(GvHD[[1]])[3:6], asinh, transformationId="asinh")
add(wf, tf)
wf
views(wf)
@ 
%
There are several things to note here. First, we didn't assign the
return value of \Rfunction{add} back to \Robject{wf}. This is the
aforementioned feature of the reference semantic. Internally,
\Rclass{workFlows} are to the most part \Rclass{environments}, and all
modifications change the content of these \Rclass{environments} rather
than the object itself. We also used the \Rfunction{add} methods with
two arguments only: the \Rclass{workFlow} and the
\Rclass{transformList}. In a hierarchical work flow we usually want to
control to which subset (or rather to which \Rclass{view}, to stick to
the work flow lingo) we want to add the \Rclass{actionItem} to. The
default is to use the base view, and alternative parent \Rclass{views}
have to be specified using the \Robject{parent} argument;
\Robject{add(wf,tf,parent="base view")} would have been
equivalent. Adding the transformation created a new \Rclass{view}
called ``asinh'' and we can see that this name was automatically taken
from the \Rclass{transformationList} object. If we want to control the
name of the new \Rclass{view}, we can do so using the \Robject{name} argument:
<<itemName>>=
add(wf, tf, name="another asinh transform")
wf
@
%
This operation created a warning since we didn't use a leaf node to
add the transformation. The software can't be sure if any of the
downstream leafs in the work flow hierarchy are affected by the
transformation (as it would certainly be true for most gates). One
solution would be to update all downstream leafs in the work flow,
but this feature is not supported yet. Also note that we used the same
\Rclass{transformList} object twice. While we controlled the name that
was used for the second \Rclass{view}, the alias for the
\Rclass{actionItem} is the same in both cases, and we would have to
use the internal unique identifier to unambiguously access the
respective object.

Transformation operations (and also compensations or normalizations)
only create single \Rclass{views}. This is not true for gating
operations. Even the most simple gate creates two populations: those
cells in the gate and the complement. Accordingly, two separate
\Rclass{views} have to be created as well. 
<<addGate>>=
rg <- rectangleGate("FSC-H"=c(200,400), "SSC-H"=c(250, 400),
                    filterId="rectangle")
add(wf, rg, parent="asinh")
wf
@
%
As we see, there are now two new \Rclass{views} under the ``asinh''
node called ``rectangle+'' and ``rectangle-''. Other filter types may
create more than two populations, for instance a \Rclass{quadrantGate}
always results in four sub-populations. There is no restriction on the
number of populations that are allowed to be created by a single
gating operation, however, when gating a \Rclass{flowSet}, the number
of populations have to be the same for each frame in the set.
<<addQuadGate>>=
qg <- quadGate("FL1-H"=2, "FL2-H"=4)
add(wf,qg,parent="rectangle+")
wf
@ 
%
For complex work flows it becomes increasingly hard to convey all
information by printing on the screen. There is a plotting function
for \Rclass{workFlow} objects which plots the underlying tree. The
function depends on the \Rpackage{Rgraphviz} package.
<<plotwf, fig=true>>=
plot(wf)
@

\subsection{Accessing items in the \Rclass{workFlow} object}
The main reason for having the \Rclass{workFlow} class is to be able
to easily access all elements of a potentially complex analysis work
flow. The predominant use case here is to access individual
\Rclass{views}, either for plotting, to create summary statistics or
to use the underlying data for subsequent analysis steps. You can do
that using the familiar list subsetting syntax:
<<getView>>=
wf[["rectangle+"]]
wf$asinh
@
%
This also works for \Rclass{actionItems}:
<<getAction>>=
wf[["action_rectangle"]]
@ 
% 
In order to retreive the underlying data for a \Rclass{view} you
can use the \Rfunction{Data} method.
<<getData>>=
Data(wf[["rectangle+"]])
@ 
%
There are \Rfunction{summary} methods defined for the different
\Rclass{view} subclasses, which basically create the respective
summaries of the associated \Rpackage{flowCore} objects, e.g. a
\Rclass{filterResult} for a filtering operation.
<<summaries>>=
summary(wf[["action_rectangle"]])
summary(wf[["CD15 FITC+CD45 PE+"]])
@ 
The \Rpackage{flowViz} package also defines \Rfunction{xyplot} and
\Rfunction{densityplot} methods for \Rclass{view} objects. Sensible defaults
are usually chosen automatically, for instance to add the gate
boundaries to the plot of a \Rclass{gateView}.
<<plotViews>>=
densityplot(wf[["base view"]])
xyplot(`FL1-H` ~ `FL2-H`, wf[["CD15 FITC+CD45 PE+"]])
@

\subsection{Removing items from \Rclass{workFlow} object}
Still need to document Rm, undo, etc.
